% Declarations for Front Matter

% MS Thesis = 0, Phd Dissertation = 1
\isdissertation{1}

% electronic submission? Paper only = 0, Electronic = 1
\iselectronic{1}

\title{Performance of a Novel Virtual Environment for Cockpit Evaluation}
\author{Richard D. Joyce}

% Choices are September, December, March, June
\degreemonth{March}
\degreeyear{2018}
%More examples DOCTOR OF PHILOSOPHY
\degree{Doctor of Philosophy}

\chair{Stephen K. Robinson}
\othermembers{Ron A. Hess,Michael Feary} %comma separated list of committee not including chair
\numberofmembers{3} % size of committee

\prevdegrees{M.S.,~University of California,~Davis, 2013\\
B.S.,~Columbia University,~2009}

%Your Graduate Group
\field{Mechanical and Aerospace Engineering}
\campus{Davis}


% add the abstract here
\input{abstract}

%Not required for electronic submission, you will need to print your other abstract page 2x and hand them in.
% Here is the first, external, abstract.
%\begin{abstract}
%	\myabstract
%	\abstractsignature
%\end{abstract}

\begin{frontmatter}
\maketitle

% A copyright page is optional. If you have one, it must immediately
% follow the title page. For more information about the copyright page
% see the UCD's Office of Graduate Studies web site.
% \copyrightpage

% dedication (optional), remove comment markers to use
%\begin{dedication}
%\null\vfil
%{\large
%\begin{center}
%xxxx
%\end{center}}
%\vfil\null
%\end{dedication}



\tableofcontents
\listoffigures
\listoftables

% Here is the second, internal, abstract.
% Update: Melissa Danforth 2006
% Inline abstract is now part of front matter according to coordinator
    \newpage
    \begin{inlineabstract}
		%Only enable small if you're trying to make it fit.
		%\begin{small}
		\myabstract
		%\end{small}
    \end{inlineabstract}

%Acknowledgments (optional)
\begin{acknowledgments}
    Thank you to Professor Robinson, who provided the inspiration for this project and has guided me throughout the whole process.
    Thank you to my committee members, Professor Hess and Dr.\ Feary, who provided support in shaping the research, especially in experimental design.
    Thank you to my fellow HRVIP lab members, who elevated this work with their high standards.
    Thank you to John Karasinski, who was always willing to discuss research over coffee.
    Thank you to Vern Belos, who guided me to finish line.
    Thank you to the Link Foundation, who selected me for the Advanced Training and Simulation Fellowship and provided financial support.
    Thank you to the San Jose State University Research Foundation, who provided financial support.
    Thank you to the subjects who volunteered their time and made this research possible.

    Thank you to my brothers, Dr.\ Joyce and Dr.\ Joyce, who inspired this endeavour. Thank you my parents, who have supported and encouraged our academic pursuits.

    Thank you to my wife, who has done more than anyone else to support me throughout this chapter of our lives. Thank you to our dogs, who never once asked me how my research was going.

\end{acknowledgments}

%Preface (optional)
% \chapter*{Preface}
% \addcontentsline{toc}{chapter}{Preface}
% \input{preface}


\end{frontmatter}
